Allee Effect and Cultural Evolution

### 2015-05-26 manuscript/draft.tex
minor updates

\begin{abstract}
Several forms of social learning rules are guided by the amount of payoff received by social teachers adopting specific cultural variants. The rationale, is that such payoff will be correlated to the beneficial or detrimental properties of a given variant. Here we assume that the payoff directly affects the reproductive fitness of individuals, and that it is determined by the density of individuals possessing the same cultural variant. More specifically we consider a  form of density-dependence where a too small or too large population of individuals possessing the same variant will be subject to a decline in fitness, while an intermediate size is optimal. This non-linear relationship, known as Allee effect in ecology, portrays a large number of cultural variants, in particular subsistence strategies that are enhanced  by the effect of cooperation but are constrained by interference and limited resources. This paper explores the evolutionary dynamics of a two trait model where each variant exhibit Allee effect. We used a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion to suboptimal variants, long-term co-existence of multiple variants, and fixation to a single optimal or suboptimal strategy.

% XAVI: I'm confused by the use of trait and variant: shouldn't it be 'a single trait model with two variants'?
\end{abstract}

@@

A key element in all social learning strategies is the availability of some heuristics and learning bias that guides the selection of a given variant over another. This information can be based on some properties of the variant itself (\emph{content-bias}) or derived from contextual cues (\emph{context-bias}), such as the rarity or the commonness or the rareness of a trait, or some features associated with the model (i.e. the social teacher) such as prestige, success, or the similarity to the learner~\citep{henrich_mcelreath2003}. An expected consequence of \emph{context-bias} is the possibility of a feedback process, whereby the attractiveness of a given variant might change as a consequence of the learning strategy~\citep{kendal_etal_2009}. For example a strong anti-conformist bias might initially favour rare traits, but as they spread into a population their selective advantage will decline as the result of a negative feedback. In a similar, but in a opposite direction, both conformist-bias and homophily-guided learning strategies would exercise a positive feedback loop.  %citation needed?

This paper examines a group of learning strategies generally referred to as \emph{payoff-based transmission}~\citep{schlag1998,kendal_etal_2009,lake_and_crema_2012,baldini2013,kandler_and_laland_2013,crema_lake_inpress}. The core assumption shared by these models is that the probability of copying a given cultural variant is positively correlated with some payoff signal expressed by individuals possessing such a variant. Several variants of this learning strategies have been proposed and explored, including: \emph{imitate the best}, where the social learner identifies and copies the trait possessed by the individual with the highest payoff; \emph{compare means}, where the social learner compare the average payoff of different strategies and chooses the highest variant; and the \emph{proportional imitation rule}, where the probability of adopting a variant is proportional to the difference in payoff between the demonstrator and the learner~\citep{schlag1998,baldini2013}.

@@

Given the same cultural trait, the variability in the payoff signal is partly derived by the presence of interacting cultural and physical traits, as well as the environment where the trait is manifest. For example, a fisherman might evaluate the performance of a hook (the target cultural trait) using as a cue the number of fish captured (the payoff signal). The correlation between the two is partly due to the efficiency of the hook, but also by the rod, the choice of the bait, the availability of the prey species, the choice of the fishing spot, the physical properties of the fisherman, and chance. In many cases we should expect that the social teacher and learner share many of these interacting traits as well as the environment where these are displayed. In other words, cultural and ecological inheritance can generally reduce the uncertainty and the variability in the payoff-signal. There are however exceptions. One of these are contexts where the success of a variant is strongly correlated with the absolute number of individuals possessing the same trait within a population. This density dependence implies the adoption (or abandonment) of a trait can alter the expected payoff-signal, possibly generating a variety of feedback mechanism.

Several cultural traits are expected to show such density dependence. In particular, behaviour linked to the exploitation of limited resources are expected to exhibit a negative density dependence. Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded,  beneficial effect will start to decline, the payoff will become smaller, and eventually rarer suboptimal variants can be preferred (see Lake and Crema 2012 for an extensive analysis). Density dependence can also have a positive direction, whereby an increase in the number of individuals possessing a trait might have a selective advantage additional to what is already expected from its frequency. Variants that are enhanced by some form of cooperation, or more in general traits exhibiting a positive niche construction are expected to show such a pattern, known in ecology as \emph{Allee Effect}~\citep{allee1958}. For example foraging groups are known to show higher performance when the number of individuals is not too small~\citep{hill_and_hawkes_1983,janssen_and_hill_2014}, while certain activities such as anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} might require a population density beyond a certain threshold in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. Other traits that might be promoted by increased density includes language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies~\citep{van_slyke_perceived_2007}.

Here we explore the combined effect of positive and negative density-dependent selection, assuming that: 1) the social learning strategy is payoff-based transmission; and 2)  payoff is equivalent to the reproductive fitness of each individual. We will demonstrate, by exploring a two-trait difference equation, that the degree by which a population relies on social learning can strongly affect the evolutionary trajectory of the system, leading to the emergence of stable and unstable equilibria, as well as the occurrence of temporary or permanent episodes of reversion to suboptimal variants.
% XAVI not sure if social learning or allee is more important...what about: "we will demonstrate...that the impact of Allee effect on social learning can strongly affect the evolutionary trajectory of the system

@@

>>>>

\begin{abstract}
Social learning rules are often guided by the amount of payoff received by potential social teachers while expressing specific cultural traits. The assumption of the social learner is that such payoff is a proxy for inferring the qualitative features of the target trait. Here we explore the theoretical implications when payoff directly affects  reproductive fitness and its expected amount is a function of the number of individuals possessing the given trait. More specifically we consider a particular form of density-dependence where a too small or too large population of individuals possessing the same variant will be subject to a decline in fitness, while an intermediate size is optimal. This non-linear relationship, known as Allee effect, portrays a large number of behavioural traits, in particular subsistence strategies that are enhanced  by the effect of cooperation but are constrained by interference and limited resources. We  explores the evolutionary dynamics of a single trait model  with two variants, each exhibiting a different Allee effect. We used a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion, stable co-existence, and fixation to one variant.

% XAVI: I'm confused by the use of trait and variant: shouldn't it be 'a single trait model with two variants'?
% I'm using them as synonym, but I guess you have a point.

\end{abstract}

@@

A central element of all social learning strategies is the presence of some heuristics that offers some guidance in the selection of a given variant over another. This  can be based on some properties of the variant itself (\emph{content-bias}) or derived from contextual cues (\emph{context-bias}), such as the commonness or the rareness of a trait, or some features associated with the transmitter such as prestige, success, or his/her similarity to the learner~\citep{henrich_mcelreath2003}. An expected consequence of \emph{context-bias} is the possibility of a feedback process, whereby the attractiveness of a given variant might change as a consequence of the learning strategy~\citep{kendal_etal_2009}. For example a strong anti-conformist bias might initially favour rare traits, but as they spread into a population their selective advantage will decline as the result of a negative feedback. On the other hand conformist-bias and homophily-guided learning strategies are expected to exercise a positive feedback to some form of fixation.

This paper examines a group of learning strategies generally referred to as \emph{payoff-based transmission}~\citep{schlag1998,kendal_etal_2009,lake_and_crema_2012,baldini2013,kandler_and_laland_2013,crema_lake_inpress}. The core assumption shared by these models is that the probability of copying a given cultural trait $i$ is positively correlated with a payoff signal $s$, determined by some function $f(i)$ expressed by individuals possessing the trait itself. Several variants of this learning strategies have been proposed and explored, including: \emph{imitate the best}, where the social learner identifies and copies the trait possessed by the individual with the highest payoff; \emph{compare means}, where the social learner compare the average payoff of different strategies and chooses the highest variant; and the \emph{proportional imitation rule}, where the probability of adopting a variant is proportional to the difference in payoff between the demonstrator and the learner~\citep{schlag1998,baldini2013}.

@@

Given the same cultural trait, the variability in the payoff signal is partly derived by the presence of interacting cultural and physical traits, as well as the environment where the trait is manifest. Thus the payoff is better described by the function $f(i,\epsilon)$, where $i$ is the trait, and $\epsilon$ is a series of additional hidden or latent variables that might or might interact with $i$. For example, a fisherman might evaluate the performance of a hook (the target cultural trait) using as a cue the number of fish captured (the payoff signal). The correlation between the two is partly due to the efficiency of the hook, but also by the rod, the choice of the bait, the availability of the prey species, the choice of the fishing spot, the physical properties of the fisherman, and chance. In many cases we should expect that the social teacher and learner share many of these interacting traits as well as the environment where these are displayed. In other words, cultural and ecological inheritance can generally reduce the impact of $\epsilon$, and consequently the uncertainty and the variability in the payoff-signal. There are however exceptions. One of these are contexts where the success of a variant is strongly correlated with the absolute number of individuals $n_i$ possessing the same trait within a population. This density dependence, which can be described by the payoff function $f(i,n_i)$, implies that the adoption (or abandonment) of a trait can alter the expected payoff-signal, possibly generating a variety of feedback mechanism.

Several cultural traits are expected to show such density dependence. In particular, behaviour linked to the exploitation of limited resources are expected to exhibit a negative density dependence. Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded (i.e. the carrying capacity),  beneficial effect will start to decline, the payoff will become smaller, and eventually rarer suboptimal variants can be preferred (see Lake and Crema 2012 for an extensive analysis).

However, density dependence does not necessarily imply only a negative correlation between population size and payoff/fitness. In many ecological contexts an inverse density dependence, whereby per capita growth rate is positively correlated with population density, has been noted. This phenomena, known as Allee effect \citep{allee1958,courchamp_etal_1999}, is generally explained by a range of processes from cooperation to unconscious mutual facilitation that enhances individual fitness at low population density. This inverse density dependence can be further distinguished between \emph{weak} and \emph{strong} Allee effect. In the former case, the reproductive fitness is always positive at low population density, whilst in the latter case we expect the presence of a critical density below which fitness becomes negative \citep{Wang_and_Kot_2001}.

Variants that are enhanced by some form of cooperation, or more in general traits exhibiting a positive niche construction are expected to show such a pattern, known in ecology as \emph{Allee Effect}~\citep{allee1958}.

While a systematic review of the empirical evidence of Allee effect has been conducted in population ecology \citep{kramer_etal_2009}, no

For example foraging groups are known to show higher performance when the number of individuals is not too small~\citep{hill_and_hawkes_1983,janssen_and_hill_2014}, while certain activities such as anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} might require a population density beyond a certain threshold in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. Other traits that might be promoted by increased density includes language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies~\citep{van_slyke_perceived_2007}.

%We definitely need to add some literature review on the theorethical models of Allee effect here as an additional paragraph.

Here we explore the combined effect of positive and negative density-dependent selection, assuming that: 1) the social learning strategy is payoff-based transmission; and 2)  payoff is equivalent to the reproductive fitness of each individual. We will demonstrate, by exploring a two-trait difference equation, that the impact of Allee effect on social learning can strongly affect the evolutionary trajectory of the system leading to the emergence of stable and unstable equilibria, as well as the occurrence of temporary or permanent episodes of reversion to suboptimal variants.

@@

%Mention that we need more empirical evidence of Allee effect
